---
output:
  html_document: default
  pdf_document: default
---
Analysis of US loan data (2005-2014) by Christopher Buss
========================================================

```{r echo=FALSE, message=FALSE, warning=FALSE, packages}
library(ggplot2)
library(dplyr)
library(GGally)
library(plyr)
library(corrplot)
library(memisc)
library(car)
library(RColorBrewer)
```

```{r echo=FALSE, Load_the_Data}
# Set path
setwd("C:/Users/chrmaier/Box Sync/04 Data Science/01 Udacity/05 EDA/03 Project")

# Import data
df <- read.csv("prosperLoanData.csv") # read csv file

# Keep original df for comparison
df_full <- df
```

# Univariate Plots Section
There are two main reasons for me to use the prosper loan data. 
First, because it is a highly relevant societal topic as we have seen how the dept 
crisis in the US in 2008 lead to a global recession. Second, because I have recently 
bougt a flat and I am interested which factors determine interest rates for individual 
customers. I begin the exploration with an univariate analysis of the dataset.

## Loans
```{r echo=FALSE, warning=FALSE}
# Recode quarters to order them
df$Quarter <- car::recode(df$LoanOriginationQuarter, "
      'Q4 2005'='05/4';      
      'Q1 2006'='06/1';
      'Q2 2006'='06/2'; 
      'Q3 2006'='06/3';
      'Q4 2006'='06/4';
      'Q1 2007'='07/1';
      'Q2 2007'='07/2'; 
      'Q3 2007'='07/3';
      'Q4 2007'='07/4';
      'Q1 2008'='08/1';
      'Q2 2008'='08/2'; 
      'Q3 2008'='08/3';
      'Q4 2008'='08/4'; 
      'Q1 2009'='09/1';
      'Q2 2009'='09/2'; 
      'Q3 2009'='09/3';
      'Q4 2009'='09/4'; 
      'Q1 2010'='10/1';
      'Q2 2010'='10/2'; 
      'Q3 2010'='10/3';
      'Q4 2010'='10/4'; 
      'Q1 2011'='11/1';
      'Q2 2011'='11/2'; 
      'Q3 2011'='11/3';
      'Q4 2011'='11/4'; 
      'Q1 2012'='12/1';
      'Q2 2012'='12/2'; 
      'Q3 2012'='12/3';
      'Q4 2012'='12/4'; 
      'Q1 2013'='13/1';
      'Q2 2013'='13/2'; 
      'Q3 2013'='13/3';
      'Q4 2013'='13/4'; 
      'Q1 2014'='14/1';
      'Q2 2014'='14/2'; 
      'Q3 2014'='14/3';
      'Q4 2014'='14/4'")

# Workaround: Counting N over quarter and saving to df  
df$count <- 1
dfCount <- aggregate(df$count, by=list(Category=df$Quarter), FUN=sum)
# Define breaks for discrete x-scale
brks = c("05/4", "06/4", "07/4", "08/4", "09/4", "10/4", "11/4", "12/4", "13/4", "14,4")

ggplot(dfCount, aes(x = Category, y = x)) +
  geom_point() +
  xlab("Quarter") +
  ylab("Number of loans") +
  ggtitle("Number of loans over time") +
  scale_x_discrete(breaks = brks)
```

The number of loans fluctuates over time. There is a steep increase towards the
end of the period of investigation.

## BorrowerRate
```{r echo=FALSE, warning=FALSE}
# Hist and summary statistics of BorrowerRate
summary(df$BorrowerRate)
ggplot(aes(x = BorrowerRate), data = df) +
  geom_histogram(breaks=seq(0,0.4, by=0.005))

"Modus:" 
names(sort(-table(df$BorrowerRate)))[1]
```

The mean interest rate for all loans is about 19%. The distribution approximates
a normal distribution with a high peak at exactly 31.77%. This might be a problem
with the data or could be a result of a special offer for loans at this rate.

```{r echo=FALSE, warning=FALSE}
# BorrowerRate over time
ggplot(aes(x = Quarter, 
          y = BorrowerRate, group = 1), 
          data = df) +
  geom_line(stat = "summary", 
      fun.y = median,
      color = "red") +
  coord_cartesian(ylim = c(0, 0.3)) +
  labs(x= "Quarter", y= "Median BorrowerRate") +
  scale_x_discrete(breaks = brks)
```

Overall, the median rate is surprisingly stable over time. However, there is an 
increase in the second half of 2010 and a decrease on the previous level from 2012
onwards.

## Prosper Score
```{r echo=FALSE, warning=FALSE}

summary(df$ProsperScore)

ggplot(aes(x = ProsperScore), data = df) +
  geom_histogram(breaks=seq(0,11, by=1))
```

## Amount
```{r echo=FALSE}
summary(df$LoanOriginalAmount)

ggplot(aes(x = LoanOriginalAmount), data = df) +
  geom_histogram(breaks=seq(0,35000, by=1000))
```

The amount of the loans ranges from \$1,000 to \$35,000 with a median loan of 
\$6,500. There are several suspisious peaks in the distribution. 

```{r echo=FALSE}
ggplot(aes(x = LoanOriginalAmount), data = df) +
  geom_histogram(breaks=seq(1,5000, by=100))

df$AmountLog <- log(df$LoanOriginalAmount)
ggplot(aes(x = AmountLog), data = df) +
  geom_histogram(breaks=seq(7,11, by=0.3))
```

Zooming in, I see that these peaks represent multiplikates of \$500 and \$1000. 
There are peaks in the distribution at \$4,000, \$10,000, \$15,000, \$20,000 
and \$25,000. Because of the skewed distribution I decide to use a log scale 
for this variable, leading to a distribution that is closer to a normal distribution.

## Monthly income 
```{r echo=FALSE, warning=FALSE}
# Rename income to make it more readable
df <- rename(df, c("StatedMonthlyIncome"="Income"))

summary(df$Income)

ggplot(df, aes(y = Income, x = "")) + 
  geom_boxplot()

ggplot(aes(x = Income), data = df) +
  geom_histogram(breaks=seq(0,20000, by=500))
```

The Boxplot reveals that there is one extreme outlier with \$1,750,003 monthly 
income, a yearly income of moren than \$21 Million. It is not impossible, but 
highly unlikely that someone with such a high income needs a loan of a few thousand 
Dollars. Therefore, I assume an error in the data and exclude this case from  
further analyses.

Because of the skewed distribution, I  decide to use a 
log scale for the income variable.

```{r echo=TRUE, warning=FALSE}
# Excluding cases with unrealistic income
df <- subset(df, Income < 1750000)

# Transform income to log-scale
df$IncomeLog <- log(df$Income+1)
ggplot(aes(x = IncomeLog), data = df) +
  geom_histogram(breaks=seq(6,11, by=0.1))
```

## MonthlyLoanPayment
```{r echo=FALSE, warning=FALSE, MonthlyPayment}
summary(df$MonthlyLoanPayment)

ggplot(aes(x = MonthlyLoanPayment), data = df) +
  geom_histogram(breaks=seq(0,2300, by=100))

ggplot(aes(x = MonthlyLoanPayment), data = df) +
  geom_histogram(breaks=seq(0,1000, by=10))
```

## Monthly payment as share of total amount
```{r echo=FALSE, warning=FALSE}
df$LoanPaymentPerc <- df$LoanOriginalAmount/df$MonthlyLoanPayment
summary(subset(df, is.finite(LoanPaymentPerc), select=LoanPaymentPerc))

ggplot(df, aes(y = LoanPaymentPerc, x = "")) + 
  geom_boxplot()

df <- subset(df, LoanPaymentPerc < 100)

ggplot(aes(x = LoanPaymentPerc), data = df) +
  geom_histogram(breaks=seq(0,100, by=1))

ggplot(aes(x = LoanPaymentPerc), data = df) +
  geom_histogram(breaks=seq(20,50, by=0.5)) +
  xlab("% of total loan that is paid back monthly")
```

I calculate the share of the monthly payments from the total loan. The majority
of payments are about 20%-40% of the total loan. There are some extreme outliers
with payments much higher than the total loan. I exclude these cases from the analysis
because these do not seem to be valid values to me.

## Term
```{r echo=FALSE, warning=FALSE}
summary(df$Term)
#table(df$Term)

# Recode "Term" to categorial variable
df$TermRe <- df$Term
df$TermRe[df$Term==12] <- 0
df$TermRe[df$Term==36] <- 1
df$TermRe[df$Term==60] <- 2
df$TermRe <- factor(df$TermRe, levels = c(0,1,2), labels = c("Short", "Medium", "Long"))
table(df$TermRe)
```

There are only 12, 36 and 60 month values in *term*. 
Therefore, I recode it to a categorial variable.

## Employment status
```{r echo=FALSE, warning=FALSE}
# Ordered Bar Chart and Table
# (https://stackoverflow.com/questions/5208679/order-bars-in-ggplot2-bar-graph?newreg=6ddc9ae45899482aaff51fbd1b098567)
ggplot(df, aes(x=reorder(EmploymentStatus,EmploymentStatus, function(x)-length(x)))) +
  geom_bar(stat="count", width=0.7, fill="steelblue") +
  xlab('Employment status')
  
table(df$EmploymentStatus)
```

The categories of the employment status variable are not mutually exclusive and
the missings are coded in different ways. Therefore, I recode the variable. Because
I do not want to exclude all missings on this variable in a later regression, I do
not mark these cases as "NA" but as "unknown".

```{r echo=TRUE, warning=FALSE}
# Recode employment
df$Employment_re<-car::recode(df$EmploymentStatus,
    "c('Employed', 'Full-time') = 'Employed';
    c('Other', 'Not available', '') = 'Unknown' ")
```

```{r echo=FALSE, warning=FALSE}
ggplot(df, aes(x=reorder(Employment_re,Employment_re, function(x)-length(x)))) +
  geom_bar(stat="count", width=0.7, fill="steelblue") +
  xlab('Employment status')
  
#table(df$Employment_re)
```

## Homeowners
```{r echo=FALSE, warning=FALSE, Cat}
"Percentage of homeowners:"
round(prop.table(table(df$IsBorrowerHomeowner)),3)
table(df$BorrowerState)
```

About half of the borrowers are home owners. They come from all 51 US states.

# Loan status
```{r echo=FALSE, warning=FALSE}
table(df$LoanStatus)

# Recode status of the loan in fewer categories
df$LoanStatusRe <- car::recode(df$LoanStatus,
    "c('Chargedoff', 'Defaulted', 'Cancelled') = 'Def/Cancelled';
    c('Past Due (>120 days)', 
    'Past Due (1-15 days)',
    'Past Due (16-30 days)',
    'Past Due (31-60 days)',
    'Past Due (61-90 days)',
    'Past Due (91-120 days)') = 'Past Due';
    'FinalPaymentInProgress' = 'Completed' ")

table(df$LoanStatusRe)
```

I recode the loan status to four categories.

## DebtToIncomeRatio
```{r echo=FALSE, warning=FALSE}
# DebtToIncomeRatio
summary(df$DebtToIncomeRatio)

ggplot(aes(y = DebtToIncomeRatio, x=""), data = df) +
  geom_boxplot()

ggplot(aes(x = DebtToIncomeRatio), data = df) +
  geom_histogram(breaks=seq(0,1, by=0.01))
```

Most borrowers have a *DebtToIncomeRatio* of about 0 to 0.75. However, there is
a number of extreme outliers up to 10. Either there are errors in the data, or
some people are highly in dept. We should take this distribution into account when
interpreting regression results later on.

## Recommendations, InvestmentsFromFriends
```{r echo=FALSE, warning=FALSE}
summary(df$Recommendations)
summary(df$InvestmentFromFriendsCount)
```

Only few borrowers have a value other than "0" for recommendations and
investments from friends. Because of this low information value, I will not 
consider these variables in the following analysis.

# Univariate Analysis

### What is the structure of your dataset?
There are 113,937 observations and 82 variables.

### What is/are the main feature(s) of interest in your dataset?
The main feature of interest is the borrower rate. I suspect that certain aspects
of the borrower and the loan itself are related to the borrower rate.

### What other features in the dataset do you think will help support your investigation?
In particular, I assume the following features are useful for my predictive model:
*Aspects of the borrower**
*	Prosper scores
*	Income
*	Dept to income ratio

*Aspects of the loan**
*	Amount
*	Monthly payment relative to loan
* Term of the loan

I also assume that other factors such as the recommendations might be related
to the interest rate. However, these variables contain little variance and therefore
are left out of the further analysis. Occupation could also be interesting, but I would
need to combine the many categories in a sensible way, which goes beyond the scope 
of this EDA.

### Did you create any new variables from existing variables in the dataset?
I calculate the share of the monthly payments from the total loan.
I recoded the variable for *employment status* by combining categories.
Also, I recoded *term* from a continous to a categorial variable.

### Of the features you investigated, were there any unusual distributions?
*Income* and the *total amount of the loan* exhibited skewed distributions. Therefore,
I calculated the logs for these values.

# Bivariate Plots Section

```{r echo=FALSE, warning=FALSE}
# Subset relevant variables
df_subset = df[, c('BorrowerRate',
          'MonthlyLoanPayment',
          'ProsperScore',
          'LoanOriginalAmount',
          'DebtToIncomeRatio',
          'IncomeLog',
          'LoanPaymentPerc')]

# Correlation matrix + visualization
cor_mat <- cor(df_subset, method = "pearson", use="complete.obs")
#round(cor_mat, 2)

# Create correlation plot
corrplot(cor_mat, type = "upper", order = "hclust",
         tl.col = "black", tl.srt = 45)

# Scatterplot matrix
set.seed(2000)
ggpairs(df_subset[sample.int(nrow(df_subset), 200),]) +
    ggtitle("Scatterplot matrix of variables of interest")
```

These plots show the correlations between the variables of interest. I proceed
with a closer look at variables that show a strong relationship or unusual patterns.

## Scatterplots - Borrower rate and independent variables
```{r echo=FALSE, warning=FALSE}
ggplot(aes(x = ProsperScore, y = BorrowerRate), data = df) +
  geom_jitter(alpha = 1/40) +                                   # add shades and noise
  geom_line(stat = "summary",                                   # add line for mean
      fun.y = mean,
      color = "yellow",
      size = 2) +
  coord_cartesian(ylim = c(0, 0.35)) +                          # Restrict y-range
  ggtitle("Borrower rate X Prosper score")

## Scatter - BorrowerRate X LoanOriginalAmount
ggplot(aes(x = BorrowerRate, y = LoanPaymentPerc), data = df) +
  geom_jitter(alpha = 1/10) +
  xlim(0,0.35) +
  ggtitle("Borrower rate X Amount")
```

*ProsperScore* is clearly related to *BorrowerRate*. Individuals with the worst
score pay amout three times more interest rates than individuals with the best
score. There is an interesting break in the line between scores 5 and 6, showing
that there is no difference in the Borrower rate between these groups.

There is a very interesting pattern for the second plot. There is a negative relationship
between the *monthly loan payment* (as share of the loan) and *borrower rate*. However,
there are different trajectories in the data. This needs more exploration in the
multivariate part.

## Scatterplots - independent variables
```{r echo=FALSE, warning=FALSE}
ggplot(aes(x = ProsperScore, y = LoanOriginalAmount), data = df) +
  geom_jitter(alpha = 1/40) +                                   # add shades and noise
  geom_line(stat = "summary",                                   # add line for mean
      fun.y = mean,
      color = "yellow",
      size = 2) +
  ggtitle("ProsperScore X LoanOriginalAmount")

ggplot(aes(x = IncomeLog, y = MonthlyLoanPayment), data = df) +
  geom_jitter(alpha = 1/50) +
  ylim(0,1000) +
  ggtitle("Income X Amount of loan")
```

There are also interesting patterns in regard to the independent variables.

## Loan status over time
```{r echo=FALSE, warning=FALSE}
# Read loan status per quarter into table
TableStatus <- round(prop.table(table(df$Quarter, df$LoanStatusRe),1),3)

# Plot this table (converted to df)
ggplot(aes(x = Var1, y = Freq, group=Var2), data = as.data.frame(TableStatus)) +
    geom_line(aes(color=Var2), size=1.2) +
    xlab("Quarter") + ylab("Proportion of all loans") +
    labs(color = "Loan status") +
    scale_x_discrete(breaks = brks)
```

The status of the loans is related to the quarter in which the loan originated.
From the earlier quarters, all loans are either completed, defaulted on or cancelled.
This reflects the short loan period in the data set. There is a surprisingly high
share of loans that defaulted, in particular in the years 2007 and 2008.
This is possibly related to the economic crisis in these years.

# Bivariate Analysis

### Talk about some of the relationships you observed in this part
Overall, the choosen variables in the data set are highly correlated. Most interesting
for the aim of this project are the correlations between *BorrowerRate* and the other
variables. There is a negative relationship between *BorrowerRate* and monthly payment
(both in absolute terms and as share of the loan), the amount of the loan, and prosper score.

### Did you observe any interesting relationships between the other features \ (not the main feature(s) of interest)?
The independent variables are also correlated with each other. For example, there
is a positive relationship between *personal income* on the one hand, and the *amount*,
*monthly payment* and *prosper score* on the other hand. Individuals with a high income
recieve bigger loans with better conditions than individuals with little income.

### What was the strongest relationship you found?
The strongest relationship is between *BorrowerRate* and *ProsperScore*.

## Multivariate Plots Section
```{r echo=FALSE, warning=FALSE}
### Loan status
ggplot(aes(x = BorrowerRate, y = LoanPaymentPerc), data = df) +
  geom_jitter(alpha = 1/100) +
  geom_point(aes(colour = factor(LoanStatusRe))) +
  xlim(0,0.35) +
  scale_colour_brewer(palette = "Dark2") +
  labs(color = "Loan status")

### Quarter
ggplot(aes(x = BorrowerRate, y = LoanPaymentPerc), data = df) +
  geom_jitter(alpha = 1/100) +
  geom_point(aes(colour = factor(Quarter))) +
  xlim(0,0.35) +
  labs(color = "Quarter")

ggplot(df, aes(BorrowerRate, colour = LoanStatusRe)) +
  geom_density(size = 1) +
  scale_colour_brewer(palette = "Dark2") +
  labs(color = "Loan status")
```

The interesting relationship between *BorrowerRate* and *LoanPaymentPerc* seem to be
related to either the status of the loan or the quarter. Those data points on the
"lines" in the plot are predominantly loans that are recent (2011 onwards) and current.
Older, completed loans often diverge from this pattern.

```{r echo=FALSE, warning=FALSE}
ggplot(aes(x = BorrowerRate, y = LoanPaymentPerc), data = df) +
  geom_jitter(alpha = 1/100) +
  geom_point(aes(colour = factor(ProsperScore))) +
  xlim(0,0.35) +
  ggtitle("Relationship between interest rate, monthly payment and score") +
  xlab("Interest rate") +
  ylab("Monthly payment relative to total amount of loan") +
  labs(color = "Score")
```

Apparently, *ProsperScores* are also related to the other two variables of interest.
For all individuals with a score assigned, the monthly payment describes perfectly
the interest rate. For individuals without a score (=earlier loans), there is much
more dispersion.

```{r echo=FALSE, warning=FALSE}
ggplot(df, aes(BorrowerRate, colour = TermRe)) +
  geom_density(size = 1) +
  labs(color = "Length of loan")

ggplot(df, aes(x=TermRe, y=BorrowerRate)) + 
  geom_boxplot()
```

Another interesting pattern emerges when plotting the distribution of *BorrowerRate*
across the legth of the loan. Borrower rates increase with term length, but the
rates of medium-length loans are more dispersed.

## Linear regression analysis
```{r echo=FALSE, warning=FALSE}
# Stepwise multivariate regressions
m1 <- lm(df$BorrowerRate ~ df$ProsperScore)
m2 <- update(m1, ~ . + df$AmountLog)
m3 <- update(m2, ~ . + df$IncomeLog)
m4 <- update(m3, ~ . + df$LoanPaymentPerc)
m5 <- update(m4, ~ . + df$TermRe)
m6 <- update(m5, ~ . + df$DebtToIncomeRatio)

# Print tables
mtable(m1, m2, m3)
mtable(m4, m5, m6)

```

In the last step I conduct a multivariate regression analysis with *BorrowerRate*
as dependent variable. I include independent variables stepwise in the model.
*ProsperScore* expains about 42% of the DV's variance. With each more Scoring-Point,
*BorrowerRate* is about 0.2 percentage points lower. Judged by R², this is the most
important predictor of interest rates in the data.
From the other variables, the length of the loan (TermRe) is also quite important.
About 30% additional explained variance can be attributed to this variable. Overall,
with an R² of about 0.83 the model fits the data quite well.

# Multivariate Analysis

### Talk about some of the relationships you observed in this part \
Were there any interesting or surprising interactions between features?

There is an interesting relationship between *ProsperScore*, *LoanPaymentPerc* and
*BorrowerRate*. For the most recent loans, *LoanPaymentPerc* is a very good predictor
for *BorrowerRate*. For older loans, the relationship us much weaker.

### Did you create any models with your dataset? Discuss limitations!
I calculated a linear regression model with the *interest rate* as dependent variable, 
and *ProsperScore*, *AmountLog*, *IncomeLog*, *LoanPaymentPerc*, *TermRe* and 
*DebtToIncomeRatio* as independent variables.

Some other variables in the dataset related to the credit history of the person
could be of interest. Another limitation is that these results indicate correlation,
but not causation.

------

# Final Plots and Summary

### Plot One
```{r echo=FALSE, warning=FALSE, Plot_One}
set.seed(2000)
ggpairs(df_subset[sample.int(nrow(df_subset), 200),]) +
    ggtitle("Scatterplot matrix of variables of interest")
```

### Description One
This matrix shows the distributions and correlations of the most important variables
in the analysis. The main variable of interest, the borrower's interest rate (*BorrowerRate*),
is related to several other variables in the dataset. The most important predictor seem to be
the borrower's score (*ProsperScore*).

### Plot Two
```{r echo=FALSE, warning=FALSE}
ggplot(aes(x = ProsperScore, y = BorrowerRate), data = df) +
  geom_jitter(alpha = 1/40) +                                   # add shades and noise
  geom_line(stat = "summary",                                   # add line for mean
      fun.y = mean,
      color = "yellow",
      size = 2) +
  coord_cartesian(ylim = c(0, 0.35)) +                          # Restrict y-range
  ggtitle("Borrower rate X Prosper score") +
  ylab("Interest rate")
```

### Description Two
The relationship between *BorrowerRate* and *ProsperScore* is further illustrated
with this plot. While there is still some variability in the data, there is a clear
downward trend of the interest rate in case of better scores.

### Plot Three
```{r echo=FALSE, warning=FALSE}
ggplot(aes(x = BorrowerRate, y = LoanPaymentPerc), data = df) +
  geom_jitter(alpha = 1/100) +
  geom_point(aes(colour = factor(ProsperScore))) +
  xlim(0,0.35) +
  ggtitle("Relationship between interest rate, monthly payment and score") +
  xlab("Interest rate") +
  ylab("Monthly payment relative to total amount of loan") +
  labs(color = "Score")
```

### Description Three
The third plot visualizes the relationship between interest rate, monthly payment 
and score. For all individuals with a score assigned, the monthly payment describes 
perfectly the interest rate. For individuals without a score (=earlier loans), 
there is much more dispersion.

# Reflection
Overall, the results are not surprising - the interest rate largely depends on the
credibility of the borrower and the characteristics of the loan. I am still puzzled
by Plot 3 in my summary. I cannot explain the pattern in this graph, but I assume that
the interest rate is calculated differently for old (completed) loans than for newer
(current) loans. However, this point needs further exploration.

I did struggle a bit with the R-syntax at times because I am a genuine user of STATA.
In particular, I had to use some workarounds to make the plots over time work.
However, the graph functions in R a really nice!

Future analysis of the dataset could explore the following points:
* How can be predict if a borrower defaults on a loan?
* Are there regional differences between states?
* How can we predict someone's ProsperScore?